Cognate beginnings to bilingual lexical growth

### 2023-05-23 manuscript/_appendix/_model-diagnostics.qmd
Add AoA estimator

We used Stan [@carpenter2017stan] as the probabilistic language behind the estimation of our Bayesian models in this study, with `brms` as its R interface [@burkner2017brms]. This language implements the Markov Chain Monte-Carlo (MCMC) algorithm to explore the posterior distribution of the model. Broadly, this algorithm is used to iteratively sample the joint sampling space of the parameters to be estimated in the model, and compute, for each value sampled, its likelihood under some probability distribution previously defined. We run `r dim(model_fit_3$fit)[2]` MCMC chains, each `r dim(model_fit_3$fit)[1]` iterations long each. The correct performance of this algorithm is critical to the quality of the statistical evidence to which the outcomes of the model lead.

@@

```{r tbl-rhats-neffs}
#| label: tbl-rhats-neffs
#| tbl-cap: "Gelman-Rubin (R-hat) convergence diagnostic of the MCMC chains in the Bayesian models."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 9
#| out-width: 100%
map_df(model_rhats, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    drop_na() |> 
    ggplot(aes(.rhat)) +
    facet_wrap(~model, nrow = 1) +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    geom_histogram(binwidth = 0.001, colour = "white", fill = "grey60") +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "MCMC samples",
         title = "Gelman-Rubin (R-hat) diagnostic") +

    map_df(model_neffs, bind_rows, .id = "model") |> 
    mutate(model = paste0("Model ", str_extract(model, "[0-9]"))) |> 
    pivot_longer(-model, 
                 names_to = ".variable",
                 values_to = ".rhat") |>
    # filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na() |> 
    ggplot(aes(.rhat)) +
    facet_wrap(~model, nrow = 1) +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    geom_histogram(binwidth = 0.1, colour = "white", fill = "grey60") +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "MCMC samples",
         title = "Ratio of effective sample size") +
    plot_layout(ncol = 1) &
    plot_annotation(tag_levels = "A") &
    theme(legend.position = "none",
          panel.grid = element_blank())
```

@@

```{r}
#| label: fig-model-pairs
#| fig-cap: "Marginal distribution and bi-variate scatterplot of posterior samples for the fixed regression coefficients in Model 3."
#| fig-height: 8
#| fig-width: 10
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| warning: false
#| message: false
color_scheme_set(rep("grey60", 6))
titles <- c("Comprehension",
            "Production",
            "Age",
            "Length",
            "Exposure",
            "Cognateness",
            "Age × Exposure",
            "Age × Cognateness",
            "Exposure × Cognateness")
posterior <- as.array(model_fit_3)
np <- nuts_params(model_fit_3) 
mcmc_pairs(posterior,
           np = np, 
           pars = c("b_Intercept[1]",
                    "b_Intercept[2]",
                    "b_age_std",
                    "b_n_phon_std",
                    "b_exposure_std",
                    "b_lv_std",
                    "b_age_std:exposure_std",
                    "b_exposure_std:lv_std"),
           draws = c(1, 0),
           grid_args = list(titles = titles),
           off_diag_args = list(size = 0.5,
                                alpha = 1/10))
```

>>>>

We used Stan [@carpenter2017stan] as the probabilistic language behind the estimation of our Bayesian models in this study, with `brms` as its R interface [@burkner2017brms]. This language implements the Markov Chain Monte-Carlo (MCMC) algorithm to explore the posterior distribution of the model. Broadly, this algorithm is used to iteratively sample the joint sampling space of the parameters to be estimated in the model, and compute, for each value sampled, its likelihood under some probability distribution previously defined. We run `r dim(model_fit$fit)[2]` MCMC chains, each `r dim(model_fit$fit)[1]` iterations long each. The correct performance of this algorithm is critical to the quality of the statistical evidence to which the outcomes of the model lead.

@@

```{r tbl-rhats-neffs}
#| label: tbl-rhats-neffs
#| tbl-cap: "MCMC convergence diagnostic of all parameters in the model"
#| echo: false
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 9
#| out-width: 100%
model_convergence |> 
    ggplot(aes(.rhat)) +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    geom_histogram(binwidth = 0.001, colour = "white", fill = "grey60") +
    labs(x = "Gelman Rubin diagnostic (R-hat)",
         y = "MCMC samples",
         title = "Gelman-Rubin (R-hat) diagnostic") +

    model_convergence |> 
    ggplot(aes(.neff)) +
    geom_vline(xintercept = 1.01, linetype = "dashed") +
    geom_histogram(binwidth = 0.1, colour = "white", fill = "grey60") +
    labs(x = "Effective sample size",
         y = "MCMC samples",
         title = "Ratio of effective sample size") +

    plot_layout(ncol = 1) &
    plot_annotation(tag_levels = "A") &
    theme(legend.position = "none",
          panel.grid = element_blank())
```

@@

```{r}
#| label: fig-model-pairs
#| fig-cap: "Marginal distribution and bi-variate scatterplot of posterior samples for the fixed regression coefficients in Model 3."
#| fig-height: 8
#| fig-width: 10
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| warning: false
#| message: false
color_scheme_set(rep("grey60", 6))

titles <- c("Comprehension",
            "Production",
            "Age",
            "Length",
            "Exposure",
            "Cognateness",
            "Age × Exposure",
            "Age × Cognateness",
            "Exposure × Cognateness",
            "Age × Exposure × Cognateness")

posterior <- as.array(model_fit)
np <- nuts_params(model_fit) 
mcmc_pairs(posterior,
           np = np, 
           pars = c("b_Intercept[1]",
                    "b_Intercept[2]",
                    "b_age_std",
                    "b_n_phon_std",
                    "b_exposure_std",
                    "b_lv_std",
                    "b_age_std:exposure_std",
                    "b_exposure_std:lv_std",
                    "b_age_std:exposure_std:lv_std"),
           draws = c(1, 0),
           grid_args = list(titles = titles),
           off_diag_args = list(size = 0.5,
                                alpha = 1/10))
```

### 2023-05-23 manuscript/_appendix/_model-predictors.qmd
Add AoA estimator

@@

Multiple algorithms have been created to compute the distance between two strings of characters. The Levenshtein distance being one of them. The this algorithm calculates the minimum number of edit operations (additions, deletions, and substitutions) that one string must go through to become identical to the other string [@levenshtein1966binary] (see @eq-levenshtein). For instance, /kaza/ (*house*, in Catalan), would need to go through one substitution to become identical to its translation equivalent in Spanish /kasa/ (/z/ is replaced by /s/). Therefore, the Levenshtein distance between /kaza/ and /kasa/ is one. For an easier handling of non-ASCII characters (fairly prevalent in Catalan and Spanish, e.g., á, ü, ñ), we used phonological transcriptions in X-SAMPA format , as opposed to IPA, for its computer-friendly set of symbols [@wells1995computer].

>>>>

```{r sample-sizes}
#| label: sample-sizes
#| echo: false
#| message: false
#| warning: false
#| include: false
tar_load(c(responses, bvq_data, ))
n_obs <- nrow(responses)

sum_obs <- responses |>
    mutate(te = factor(te)) |>
    count(te) |>
    summarise(across(n, lst(median, min, max), 
                     .names = "{.fn}"))

# attrition
n_items <- bvq_data$pool

n_items_nouns <- n_items |> 
    dplyr::filter(class=="Noun")

n_items_freq <- n_items_nouns |> 
    left_join(childes,
              by = c("childes_lemma" = "token")) |> 
    drop_na(freq_zipf, wordbank_lemma)

n_items_lemma <- n_items_freq |> 
    dplyr::filter(n_lemmas == 1)

n_items_multi <- n_items_lemma |> 
    dplyr::filter(!is_multiword) |> 
    dplyr::filter(include, !is.na(wordbank_lemma)) |> 
    drop_na(xsampa) |> 
    distinct(te, language)

n_items_duplicated <- n_items_multi |> 
    add_count(te, name = "n_te") |>
    dplyr::filter(n_te == 2)

n_items_final <- n_items_duplicated |> 
    count(language) |> 
    group_split(language) |> 
    set_names(c("Catalan", "Spanish")) |>
    map_int("n")

n_te <- length(unique(responses$te))
```

@@

Multiple algorithms have been created to compute the distance between two strings of characters. The Levenshtein distance being one of them. The this algorithm calculates the minimum number of edit operations (additions, deletions, and substitutions) that one string must go through to become identical to the other string [@levenshtein1966binary] (see @eq-levenshtein). For instance, /\textipa{"ka.za}/ (*house*, in Catalan), would need to go through one substitution to become identical to its translation equivalent in Spanish /\textipa{"ka.sa}/ (/z/ is replaced by /s/). Therefore, the Levenshtein distance between /\textipa{"ka.za}/ and /\textipa{"ka.sa}/ is one. For an easier handling of non-ASCII characters (fairly prevalent in Catalan and Spanish, e.g., á, ü, ñ), we used phonological transcriptions in X-SAMPA format, as opposed to IPA, for its computer-friendly set of symbols [@wells1995computer].
